Calendar-based graphics for visualising people's daily schedules

### 2017-07-09 main.Rmd
updates

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Di Cook
    affiliation: Monash Univeristy
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash Univeristy
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  It is challenging to visualise people's daily schedules as such data are often of multiple seasons and of long time span. One example is Melbourne pedestrian sensor data which are collected at hourly interval and at a number of locations. Owing to human-related activities, the primary depiction of the data features multiple temopral scales such as time of day, day of week, day of year. We develop the \code{frame\_calendar} function that provides a method for rearranging such kind of data in a calendar-based format and accomplishing visualisation with the grammar of graphics, and describe its construction in detail. We provide some examples to demonstrate its usage by applying to the pedetrian data.
keywords:
  # at least one keyword must be supplied
  formatted: [calendar-based visualisation, temporal-context data]
  plain:     [calendar-based visualisation, temporal-context data]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. To take a glimpse at such temporal-context data, we start with the conventional time series plot (Figure \ref{fig:time-series-plot}). Figure \ref{fig:facet-time} lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

@@

```{r time-series-plot, fig.height = 10.5, fig.cap = "Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show the foot traffic varies from one sensor to another in terms of both time and number. The traffic, specifically at two train stations (Flagstaff Station and Flinders Street Station), is primarily driven by the commuters, as depicted in the graph."}
```

@@

The remainder of the paper is organised as follows. Section \ref{sec:algorithm} describes the \code{frame_calendar} function and the detailed construction. Section \ref{sec:examples} presents some examples of its usage. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

@@

In the section, the algorithms to construct calendar grids, different scales, and reference lines and labels are discussed in detail.

```
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
```

```{r flinders-2016, fig.height = 6.5, fig.cap = "The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of 3 by 4 on the monthly calendar format presents all the traffic in 2016. The disparities between weekday and weekend along with public holiday pop out to viewers in a clearer manner."}
```

@@

The algorithm can be simplified to accommodate the other two types of calendar formats. One is comprised of days of a week in columns and weeks of a year in rows, and the other is days of a month in columns and months of a year in rows, which we refer to as "weekly" and "daily" calendar respectively. Due to the arrangements, the weekly calendar puts more emphases on days of a week over days of a year, whereas the daily one serves as the opposite. The monthly format can be considered as the advanced twist of both weekly and daily. Temporal patterns lead to which format to be employed. The weekly calendar can be a nice attempt if the most variations can be characterised by days of a week. On the other hand, the absence of weekly effect but the presence of yearly effect can direct to the daily calendar. When both are present, the monthly calendar appears to be a better choice.

We only illustrate that grids are laid out over the horizontal direction in the paper. The vertical direction can be enabled by swapping $i$ and $j$ in the algorithm stated above. It is particularly useful for those users who get accustomed to calendars of vertical organisation in some countries.

@@

Section \ref{sec:grid} discusses the implementation applied to the fixed scale which is using the whole range of all the data values. In addition to the fixed scale (\code{fixed}), the remaining options include free scale on all the days (\code{free}), day of the week only (\code{free_wday}), and day of the month only (\code{free_mday}) for different comparisons.

Grouping the cells according to the common period give rise to these scales. The minimums and maximums obtained for every $(i, j)$ and $(m, n)$ allow for individual scale over all the single days. Similarly, $j$ indexing day of the week is bundled to determine the scales on days of a week, and $g$ representing the day in the month is for the scales on days of a month.

All the figures shown before are drawn on the fixed scale so that it makes the magnitudes comparable over the whole period of time. On the other hand, some sub-series of small magnitudes yet worth-noting shapes possibly become invisible whilst embedded in the overall picture. These sorts of scales, as a result, complements the calendar-based visualisation in general. (NEED FIGURES)

```{r flinders-free, fig.height = 6.5, fig.cap = "Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape becomes more distinctive at the expense of the magnitude comparison loss."}
```

@@

```{r scatterplot, fig.height = 5, fig.cap = "Lag scatterplot on the `daily' calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at the order of 1. The stronger association between the one-hour-ahead count and the present takes place on non-working days."}
```

@@

The limitation is also evident: hard to perceive trend as not on the common scale.

>>>>

---
author:
  - name: Earo Wang
    affiliation: Monash University
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{earo.wang@monash.edu}
  - name: Dianne Cook
    affiliation: Monash Univeristy
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{dicook@monash.edu}
  - name: Rob J Hyndman
    affiliation: Monash Univeristy
    address: >
      Department of Econometrics and Business Statistics,
      Monash University, VIC 3800
      Australia
    email: \email{rob.hyndman@monash.edu}
title:
  formatted: "Calendar-based graphics for visualising people's daily schedules"
  plain: "Calendar-based graphics for visualising people's daily schedules"
abstract: >
  It is challenging to visualise people's daily schedules as such data are often of multiple seasons and of long time span. One example is Melbourne pedestrian sensor data which are collected at hourly interval and at a number of locations. Owing to human activities, the data features multiple temopral scales such as time of day, day of week, day of year. We develop the \code{frame\_calendar} function that provides a method for rearranging such kind of data in a calendar-based format and accomplishing visualisation with the grammar of graphics, and describe its construction in detail. We provide some examples to demonstrate its usage by applying to the pedetrian data.
keywords:
  # at least one keyword must be supplied
  formatted: [calendar-based visualisation, temporal-context data]
  plain:     [calendar-based visualisation, temporal-context data]
preamble: >
  \usepackage{amsmath}
output: 
  rticles::jss_article:
  citation_package: biblatex
bibliography: ["references.bib", "packages.bib"]
---

@@

This work was originally motivated by studying foot traffic in the city of Melbourne [@ped]. There have been 43 installed sensors counting pedestrians every hour across the downtown area until the end of 2016 (see Figure \ref{fig:ped-map}). The dataset can shed lights into understanding people's daily schedules, or assisting administration and business planning. We start off with the conventional time series plot to take a glimpse at such data. Small multiples, shown as Figure \ref{fig:time-series-plot}, give an overall picture of foot traffic at a selection of 7 sensors. Figure \ref{fig:facet-time}, complemented to the Figure \ref{fig:time-series-plot}, provides another aspect of temporal patterns in the more detailed level by faceting day of the week. It lends itself to a number of issues that make exploratory data visualisation challenging in many temporal-context applications involving human behaviours:

@@

```{r time-series-plot, fig.height = 10.5, fig.cap = "Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The traffic, specifically at two train stations (Flagstaff Station and Flinders Street Station), is primarily driven by the commuters, as depicted in the graph."}
```

@@

The remainder of the paper is organised as follows. Section \ref{sec:algorithm} describes the \code{frame_calendar} function and its construction. Section \ref{sec:examples} presents some examples of its usage. Section \ref{sec:discussion} discusses the advantages and disadvantages of the method.

@@

Figure \ref{fig:flinders-2016} shows the line glyphs organised in the monthly calendar over the year of 2016. This is achieved by the \code{frame_calendar} function computing the new coordinates according to the input data variables; in turn the rearranged data values are plotted using the \pkg{ggplot2} package [@R-ggplot2], which is the implementation of the grammar of graphics (@wilkinson2006grammar and @wickham2010layered). The \code{frame_calendar} function assumes that the underlying data structure is tidy data [@wickham2014tidy].

<!-- ``` -->
<!-- frame_calendar( -->
<!--   data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE,  -->
<!--   nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed" -->
<!-- ) -->
<!-- ``` -->

```{r flinders-2016, fig.height = 6.5, fig.cap = "The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of 3 by 4 on the monthly calendar format presents all the traffic in 2016. The disparities between weekday and weekend along with public holiday pop out to viewers in the clearer manner."}
```

The algorithms to construct calendar grids, several scales, as well as reference lines and labels are described in the section.

@@

The algorithm can be simplified to accommodate the other two types of calendar formats. One is comprised of days of a week in columns and weeks of a year in rows, and the other is days of a month in columns and months of a year in rows, which we refer to as "weekly" and "daily" calendar respectively. Due to the way of the arrangement, the weekly calendar puts more emphases on days of a week over days of a year, whereas the daily one serves as the opposite. The monthly format can be considered as the advanced twist of both weekly and daily. Temporal patterns lead to which format to be employed. The weekly calendar can be a nice attempt if the most variations can be characterised by days of a week. On the other hand, the absence of weekly effect but the presence of yearly effect can direct to the daily calendar. When both are present, the monthly calendar appears to be a better choice.

We only illustrate that grids are laid out over the horizontal direction in the paper. The vertical direction can be enabled by swapping $i$ and $j$ in the algorithm stated above. It particularly benefits those users who get accustomed to calendars of vertical organisation in some countries.

@@

Section \ref{sec:grid} discusses the implementation applied to the fixed scale which is using the whole range of all the data values. The \code{scale} argument controls the scaling of the display. In addition to the fixed scale (\code{fixed}), the remaining options include free scale on all the days (\code{free}), day of the week only (\code{free_wday}), and day of the month only (\code{free_mday}) for different comparisons.

Grouping the cells according to the common period give rise to these scales. The minimums and maximums obtained for every $(i, j)$ and $(m, n)$ allow for individual scale over all the single days. Similarly, $j$ indexing day of the week is grouped to determine the scales on days of a week, and $g$ representing the day in the month is for the scales on days of a month.

Figure \ref{fig:flinders-2016} is drawn on the fixed scale so that it makes the magnitudes comparable over the whole period of time. On the other hand, some sub-series of small magnitudes yet worth-noting behaviours may become invisible whilst embedded in the overall picture. These sorts of scales, as a result, complements the calendar-based visualisation in general. Figure \ref{fig:flinders-free} provides an example of the free scale on all the days.

```{r flinders-free, fig.height = 6.5, fig.cap = "Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive at the expense of the magnitude comparison loss."}
```

@@

```{r scatterplot, fig.height = 5, fig.cap = "Lag scatterplot on the daily calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at the order of 1. The positive association between the one-hour-ahead count and the present regardless of the data values takes place on non-working days."}
```

@@

### 2017-07-09 main.tex
updates

\author{
Earo Wang\\Monash University \And Di Cook\\Monash Univeristy \And Rob J Hyndman\\Monash Univeristy
}
\title{Calendar-based graphics for visualising people's daily schedules}

\Plainauthor{Earo Wang, Di Cook, Rob J Hyndman}
\Plaintitle{Calendar-based graphics for visualising people's daily schedules}

\Abstract{
It is challenging to visualise people's daily schedules as such data are
often of multiple seasons and of long time span. One example is
Melbourne pedestrian sensor data which are collected at hourly interval
and at a number of locations. Owing to human-related activities, the
primary depiction of the data features multiple temopral scales such as
time of day, day of week, day of year. We develop the
\code{frame\_calendar} function that provides a method for rearranging
such kind of data in a calendar-based format and accomplishing
visualisation with the grammar of graphics, and describe its
construction in detail. We provide some examples to demonstrate its
usage by applying to the pedetrian data.
}

@@

      Di Cook\\
  Monash Univeristy\\
  Department of Econometrics and Business Statistics, Monash University,
  VIC 3800 Australia\\
  E-mail: \email{dicook@monash.edu}\\

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. To take a glimpse at such temporal-context data, we
start with the conventional time series plot (Figure
\ref{fig:time-series-plot}). Figure \ref{fig:facet-time} lends itself to
a number of issues that make exploratory data visualisation challenging
in many temporal-context applications involving human behaviours:

@@

\caption[Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne]{Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show the foot traffic varies from one sensor to another in terms of both time and number. The traffic, specifically at two train stations (Flagstaff Station and Flinders Street Station), is primarily driven by the commuters, as depicted in the graph.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

In the section, the algorithms to construct calendar grids, different
scales, and reference lines and labels are discussed in detail.

\begin{verbatim}
frame_calendar(
  data, x, y, date, calendar = "monthly", dir = "h", sunday = FALSE, 
  nrow = NULL, ncol = NULL, polar = FALSE, scale = "fixed"
)
\end{verbatim}

@@

\caption[The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs]{The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of 3 by 4 on the monthly calendar format presents all the traffic in 2016. The disparities between weekday and weekend along with public holiday pop out to viewers in a clearer manner.}\label{fig:flinders-2016}
\end{figure}
\end{CodeChunk}

@@

The algorithm can be simplified to accommodate the other two types of
calendar formats. One is comprised of days of a week in columns and
weeks of a year in rows, and the other is days of a month in columns and
months of a year in rows, which we refer to as ``weekly'' and ``daily''
calendar respectively. Due to the arrangements, the weekly calendar puts
more emphases on days of a week over days of a year, whereas the daily
one serves as the opposite. The monthly format can be considered as the
advanced twist of both weekly and daily. Temporal patterns lead to which
format to be employed. The weekly calendar can be a nice attempt if the
most variations can be characterised by days of a week. On the other
hand, the absence of weekly effect but the presence of yearly effect can
direct to the daily calendar. When both are present, the monthly
calendar appears to be a better choice.

We only illustrate that grids are laid out over the horizontal direction
in the paper. The vertical direction can be enabled by swapping \(i\)
and \(j\) in the algorithm stated above. It is particularly useful for
those users who get accustomed to calendars of vertical organisation in
some countries.

@@

Section \ref{sec:grid} discusses the implementation applied to the fixed
scale which is using the whole range of all the data values. In addition
to the fixed scale (\code{fixed}), the remaining options include free
scale on all the days (\code{free}), day of the week only
(\code{free_wday}), and day of the month only (\code{free_mday}) for
different comparisons.

Grouping the cells according to the common period give rise to these
scales. The minimums and maximums obtained for every \((i, j)\) and
\((m, n)\) allow for individual scale over all the single days.
Similarly, \(j\) indexing day of the week is bundled to determine the
scales on days of a week, and \(g\) representing the day in the month is
for the scales on days of a month.

All the figures shown before are drawn on the fixed scale so that it
makes the magnitudes comparable over the whole period of time. On the
other hand, some sub-series of small magnitudes yet worth-noting shapes
possibly become invisible whilst embedded in the overall picture. These
sorts of scales, as a result, complements the calendar-based
visualisation in general. (NEED FIGURES)

@@

\caption[Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days]{Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape becomes more distinctive at the expense of the magnitude comparison loss.}\label{fig:flinders-free}
\end{figure}
\end{CodeChunk}

@@

\caption[Lag scatterplot on the `daily' calendar]{Lag scatterplot on the `daily' calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at the order of 1. The stronger association between the one-hour-ahead count and the present takes place on non-working days.}\label{fig:scatterplot}
\end{figure}
\end{CodeChunk}

>>>>

\author{
Earo Wang\\Monash University \And Dianne Cook\\Monash Univeristy \And Rob J Hyndman\\Monash Univeristy
}
\title{Calendar-based graphics for visualising people's daily schedules}

\Plainauthor{Earo Wang, Dianne Cook, Rob J Hyndman}
\Plaintitle{Calendar-based graphics for visualising people's daily schedules}

\Abstract{
It is challenging to visualise people's daily schedules as such data are
often of multiple seasons and of long time span. One example is
Melbourne pedestrian sensor data which are collected at hourly interval
and at a number of locations. Owing to human activities, the data
features multiple temopral scales such as time of day, day of week, day
of year. We develop the \code{frame\_calendar} function that provides a
method for rearranging such kind of data in a calendar-based format and
accomplishing visualisation with the grammar of graphics, and describe
its construction in detail. We provide some examples to demonstrate its
usage by applying to the pedetrian data.
}

@@

      Dianne Cook\\
  Monash Univeristy\\
  Department of Econometrics and Business Statistics, Monash University,
  VIC 3800 Australia\\
  E-mail: \email{dicook@monash.edu}\\

@@

This work was originally motivated by studying foot traffic in the city
of Melbourne \citep{ped}. There have been 43 installed sensors counting
pedestrians every hour across the downtown area until the end of 2016
(see Figure \ref{fig:ped-map}). The dataset can shed lights into
understanding people's daily schedules, or assisting administration and
business planning. We start off with the conventional time series plot
to take a glimpse at such data. Small multiples, shown as Figure
\ref{fig:time-series-plot}, give an overall picture of foot traffic at a
selection of 7 sensors. Figure \ref{fig:facet-time}, complemented to the
Figure \ref{fig:time-series-plot}, provides another aspect of temporal
patterns in the more detailed level by faceting day of the week. It
lends itself to a number of issues that make exploratory data
visualisation challenging in many temporal-context applications
involving human behaviours:

@@

\caption[Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne]{Time series plot about the number of pedestrians over the year of 2016 at a selection of 7 sensors in the city of Melbourne. Coloured by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The traffic, specifically at two train stations (Flagstaff Station and Flinders Street Station), is primarily driven by the commuters, as depicted in the graph.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

@@

Figure \ref{fig:flinders-2016} shows the line glyphs organised in the
monthly calendar over the year of 2016. This is achieved by the
\code{frame_calendar} function computing the new coordinates according
to the input data variables; in turn the rearranged data values are
plotted using the \pkg{ggplot2} package \citep{R-ggplot2}, which is the
implementation of the grammar of graphics (\citet{wilkinson2006grammar}
and \citet{wickham2010layered}). The \code{frame_calendar} function
assumes that the underlying data structure is tidy data
\citep{wickham2014tidy}.

@@

\caption[The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs]{The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of 3 by 4 on the monthly calendar format presents all the traffic in 2016. The disparities between weekday and weekend along with public holiday pop out to viewers in the clearer manner.}\label{fig:flinders-2016}
\end{figure}
\end{CodeChunk}

The algorithms to construct calendar grids, several scales, as well as
reference lines and labels are described in the section.

@@

The algorithm can be simplified to accommodate the other two types of
calendar formats. One is comprised of days of a week in columns and
weeks of a year in rows, and the other is days of a month in columns and
months of a year in rows, which we refer to as ``weekly'' and ``daily''
calendar respectively. Due to the way of the arrangement, the weekly
calendar puts more emphases on days of a week over days of a year,
whereas the daily one serves as the opposite. The monthly format can be
considered as the advanced twist of both weekly and daily. Temporal
patterns lead to which format to be employed. The weekly calendar can be
a nice attempt if the most variations can be characterised by days of a
week. On the other hand, the absence of weekly effect but the presence
of yearly effect can direct to the daily calendar. When both are
present, the monthly calendar appears to be a better choice.

We only illustrate that grids are laid out over the horizontal direction
in the paper. The vertical direction can be enabled by swapping \(i\)
and \(j\) in the algorithm stated above. It particularly benefits those
users who get accustomed to calendars of vertical organisation in some
countries.

@@

Section \ref{sec:grid} discusses the implementation applied to the fixed
scale which is using the whole range of all the data values. The
\code{scale} argument controls the scaling of the display. In addition
to the fixed scale (\code{fixed}), the remaining options include free
scale on all the days (\code{free}), day of the week only
(\code{free_wday}), and day of the month only (\code{free_mday}) for
different comparisons.

Grouping the cells according to the common period give rise to these
scales. The minimums and maximums obtained for every \((i, j)\) and
\((m, n)\) allow for individual scale over all the single days.
Similarly, \(j\) indexing day of the week is grouped to determine the
scales on days of a week, and \(g\) representing the day in the month is
for the scales on days of a month.

Figure \ref{fig:flinders-2016} is drawn on the fixed scale so that it
makes the magnitudes comparable over the whole period of time. On the
other hand, some sub-series of small magnitudes yet worth-noting
behaviours may become invisible whilst embedded in the overall picture.
These sorts of scales, as a result, complements the calendar-based
visualisation in general. Figure \ref{fig:flinders-free} provides an
example of the free scale on all the days.

@@

\caption[Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days]{Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive at the expense of the magnitude comparison loss.}\label{fig:flinders-free}
\end{figure}
\end{CodeChunk}

@@

\caption[Lag scatterplot on the daily calendar]{Lag scatterplot on the daily calendar. Lagged hourly counts are plotted against current ones to demonstrate the autocorrelation at the order of 1. The positive association between the one-hour-ahead count and the present regardless of the data values takes place on non-working days.}\label{fig:scatterplot}
\end{figure}
\end{CodeChunk}
